---
title: "Effective Visualizations"
linktitle: "2: Visualization"
date: "2021-02-02"
read_date: "2021-02-02"
publishDate: "2021-01-19"
draft: false
menu:
  content:
    parent: Course content
    weight: 1
type: docs
output:
  blogdown::html_page:
    toc: true
---

## Ease comparisons



## Plots for two variables
In general, you should use scatterplots to visualize the relationship between two variables.
In every single instance in which we have examined the relationship between two variables, including total murders versus population size, we have used scatterplots. This is the plot we generally recommend. However, there are some exceptions and we describe two alternative plots here: the _slope chart_ and the _Bland-Altman plot_.

### Slope charts 

One exception where another type of plot may be more informative is when you are comparing variables of the same type, but at different time points and for a relatively small number of comparisons. For example, comparing life expectancy between 2010 and 2015. In this case, we might recommend a _slope chart_.

There is no geometry for slope charts in __ggplot2__, but we can construct one using `geom_line`. We need to do some tinkering to add labels. Below is an example comparing 2010 to 2015 for large western countries: 

```{r slope-plot}
west <- c("Western Europe","Northern Europe","Southern Europe",
          "Northern America","Australia and New Zealand")

dat <- gapminder %>%
  filter(year%in% c(2010, 2015) & region %in% west &
           !is.na(life_expectancy) & population > 10^7)

dat %>%
  mutate(location = ifelse(year == 2010, 1, 2),
         location = ifelse(year == 2015 &
                             country %in% c("United Kingdom", "Portugal"),
                           location+0.22, location),
         hjust = ifelse(year == 2010, 1, 0)) %>%
  mutate(year = as.factor(year)) %>%
  ggplot(aes(year, life_expectancy, group = country)) +
  geom_line(aes(color = country), show.legend = FALSE) +
  geom_text(aes(x = location, label = country, hjust = hjust),
            show.legend = FALSE) +
  xlab("") + ylab("Life Expectancy")
```

An advantage of the slope chart is that it permits us to quickly get an idea of changes based on the slope of the lines. Although we are using angle as the visual cue, we also have position to determine the exact values. Comparing the improvements is a bit harder with a scatterplot:


```{r scatter-plot-instead-of-slope, echo=FALSE}
library(ggrepel)
west <- c("Western Europe","Northern Europe","Southern Europe",
          "Northern America","Australia and New Zealand")

dat <- gapminder %>%
  filter(year%in% c(2010, 2015) & region %in% west &
           !is.na(life_expectancy) & population > 10^7)

dat %>%
  mutate(year = paste0("life_expectancy_", year)) %>%
  select(country, year, life_expectancy) %>%
  spread(year, life_expectancy) %>%
  ggplot(aes(x=life_expectancy_2010,y=life_expectancy_2015, label = country)) +
  geom_point() + geom_text_repel() +
  scale_x_continuous(limits=c(78.5, 83)) +
  scale_y_continuous(limits=c(78.5, 83)) +
  geom_abline(lty = 2) +
  xlab("2010") +
  ylab("2015")
```

In the scatterplot, we have followed the principle _use common axes_ since we are comparing these before and after. However, if we have many points, slope charts stop being useful as it becomes hard to see all the lines.

### Bland-Altman plot

Since we are primarily interested in the difference, it makes sense to dedicate one of our axes to it. The Bland-Altman plot, also known as the Tukey mean-difference plot and the MA-plot, shows the difference versus the average:

```{r, bland-altman}
library(ggrepel)
dat %>%
  mutate(year = paste0("life_expectancy_", year)) %>%
  select(country, year, life_expectancy) %>%
  spread(year, life_expectancy) %>%
  mutate(average = (life_expectancy_2015 + life_expectancy_2010)/2,
         difference = life_expectancy_2015 - life_expectancy_2010) %>%
  ggplot(aes(average, difference, label = country)) +
  geom_point() +
  geom_text_repel() +
  geom_abline(lty = 2) +
  xlab("Average of 2010 and 2015") +
  ylab("Difference between 2015 and 2010")
```

Here, by simply looking at the y-axis, we quickly see which countries have shown the most improvement. We also get an idea of the overall value from the x-axis.

## Encoding a third variable

An earlier scatterplot showed the relationship between infant survival and average income. Below is a version of this plot that encodes three variables: OPEC membership, region, and population.

```{r encoding-third-variable, echo=FALSE}
present_year <- 2010

dat <- gapminder %>%
  mutate(region = case_when(
    region %in% west ~ "The West",
    region %in% "Northern Africa" ~ "Northern Africa",
    region %in% c("Eastern Asia", "South-Eastern Asia") ~ "East Asia",
    region == "Southern Asia"~ "Southern Asia",
    region %in% c("Central America", "South America", "Caribbean") ~ "Latin America",
    continent == "Africa" & region != "Northern Africa" ~ "Sub-Saharan Africa",
    region %in% c("Melanesia", "Micronesia", "Polynesia") ~ "Pacific Islands"),
    dollars_per_day = gdp / population / 365) %>%
  filter(year %in% present_year & !is.na(gdp) & !is.na(infant_mortality) & !is.na(region) ) %>%
  mutate(OPEC = ifelse(country%in%opec, "Yes", "No"))

dat %>%
  ggplot(aes(dollars_per_day, 1 - infant_mortality/1000,
             col = region, size = population/10^6,
             pch =  OPEC)) +
  scale_x_continuous(trans = "log2", limits=c(0.25, 150)) +
  scale_y_continuous(trans = "logit",limit=c(0.875, .9981),
                     breaks=c(.85,.90,.95,.99,.995,.998)) +
  geom_point(alpha = 0.5) +
  ylab("Infant survival proportion")
```

We encode categorical variables with color and shape. These shapes can be controlled with `shape`  argument. Below are the shapes available for use in R. For the last five, the color goes inside.

```{r available-shapes, echo=FALSE, fig.height=2.25}
dat=data.frame(x=c(0:25))
ggplot() +
  theme_minimal() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
scale_shape_identity() + scale_y_reverse() +
geom_point(dat, mapping=aes(x%%9, x%/%9, shape=x), size=4, fill="blue") +
geom_text(dat, mapping=aes(x%%9, x%/%9+0.25, label=x), size=4)
```

For continuous variables, we can use color, intensity, or size. We now show an example of how we do this with a case study.

When selecting colors to quantify a numeric variable, we choose between two options: sequential and diverging. Sequential colors are suited for data that goes from high to low.  High values are clearly distinguished from low values. Here are some examples offered by the package `RColorBrewer`:

```{r eval=FALSE}
library(RColorBrewer)
display.brewer.all(type="seq")
```

```{r r-color-brewer-seq, fig.height=3.5, echo=FALSE}
library(RColorBrewer)
rafalib::mypar()
display.brewer.all(type="seq")
```

Diverging colors are used to represent values that diverge from a center. We put equal emphasis on both ends of the data range: higher than the center and lower than the center. An example of when we would use a divergent pattern would be if we were to show height in standard deviations away from the average. Here are some examples of divergent patterns:

```{r eval=FALSE}
library(RColorBrewer)
display.brewer.all(type="div")
```


```{r r-color-brewer-div, fig.height=2.5, echo=FALSE}
library(RColorBrewer)
rafalib::mypar()
display.brewer.all(type="div")
```



:::fyi

**TRY IT**

For these exercises, we will be using the vaccines data in the __dslabs__ package:

```{r}
library(dslabs)
data(us_contagious_diseases)
```

1. Pie charts are appropriate:

a. When we want to display percentages.
b. When __ggplot2__ is not available.
c. When I am in a bakery.
d. Never. Barplots and tables are always better.

2. What is the problem with the plot below:

```{r baplot-not-from-zero-exercises, echo=FALSE, message=FALSE}
library(tidyverse)
ds_theme_set()
data.frame(candidate=c("Clinton","Trump"), electoral_votes = c(232, 306)) %>%
  ggplot(aes(candidate, electoral_votes)) +
  geom_bar(stat = "identity", width=0.5, color =1, fill = c("Blue","Red")) +
  coord_cartesian(ylim=c(200,310)) +
  ylab("Electoral Votes") +
  xlab("") +
  ggtitle("Results of Presidential Election 2016")
```

a. The values are wrong. The final vote was 306 to 232.
b. The axis does not start at 0. Judging by the length, it appears Trump received 3 times as many votes when, in fact, it was about 30% more.
c. The colors should be the same.
d. Percentages should be shown as a pie chart.


3. Take a look at the following two plots. They show the same information: 1928 rates of measles across the 50 states.

```{r measels-exercise, fig.height = 5, echo=FALSE}
library(gridExtra)
p1 <- us_contagious_diseases %>%
  filter(year == 1928 & disease=="Measles" & count>0 & !is.na(population)) %>%
  mutate(rate = count / population * 10000 * 52 / weeks_reporting) %>%
  ggplot(aes(state, rate)) +
  geom_bar(stat="identity") +
  coord_flip() +
  xlab("")

p2 <- us_contagious_diseases %>%
  filter(year == 1928 & disease=="Measles" & count>0 & !is.na(population)) %>%
  mutate(rate = count / population * 10000*52 / weeks_reporting) %>%
  mutate(state = reorder(state, rate)) %>%
  ggplot(aes(state, rate)) +
  geom_bar(stat="identity") +
  coord_flip() +
  xlab("")
grid.arrange(p1, p2, ncol = 2)
```
Which plot is easier to read if you are interested in determining which are the best and worst states in terms of rates, and why?

a. They provide the same information, so they are both equally as good.
b. The plot on the right is better because it orders the states alphabetically.
c. The plot on the right is better because alphabetical order has nothing to do with the disease and by ordering according to actual rate, we quickly see the states with most and least rates.
d. Both plots should be a pie chart.


4. To make the plot on the left, we have to reorder the levels of the states' variables.

```{r}
dat <- us_contagious_diseases %>%
  filter(year == 1967 & disease=="Measles" & !is.na(population)) %>%
  mutate(rate = count / population * 10000 * 52 / weeks_reporting)
```


Note what happens when we make a barplot:

```{r barplot-plot-exercise-example, fig.height = 5}
dat %>% ggplot(aes(state, rate)) +
  geom_bar(stat="identity") +
  coord_flip()
```

Define these objects:

```{r, eval=FALSE}
state <- dat$state
rate <- dat$count/dat$population*10000*52/dat$weeks_reporting
```

Redefine the `state` object so that the levels are re-ordered. Print the new object `state` and its levels so you can see that the vector is not re-ordered by the levels.


5. Now with one line of code, define the `dat` table as done above, but change the use mutate to create a rate variable and re-order the state variable so that the levels are re-ordered by this variable. Then make a barplot using the code above, but for this new `dat`.


6. Say we are interested in comparing gun homicide rates across regions of the US. We see this plot:


```{r us-murders-barplot}
library(dslabs)
data("murders")
murders %>% mutate(rate = total/population*100000) %>%
group_by(region) %>%
summarize(avg = mean(rate)) %>%
mutate(region = factor(region)) %>%
ggplot(aes(region, avg)) +
geom_bar(stat="identity") +
ylab("Murder Rate Average")
```


and decide to move to a state in the western region. What is the main problem with this interpretation?

a. The categories are ordered alphabetically.
b. The graph does not show standarad errors.
c. It does not show all the data. We do not see the variability within a region and it's possible that the safest states are not in the West.
d. The Northeast has the lowest average.


7. Make a boxplot of the murder rates defined as

```{r, eval = FALSE}
data("murders")
murders %>% mutate(rate = total/population*100000)
```

by region, showing all the points and ordering the regions by their median rate.


8. The plots below show three continuous variables.

```{r pseudo-3d-exercise, fig.width=7, fig.height = 3.708,  echo=FALSE}
library(scatterplot3d)
library(RColorBrewer)
set.seed(1)
n <- 25
group <- rep(1,n)
group[1:(round(n/2))] <- 2
x <- rnorm(n, group, .33)
y <- rnorm(n, group, .33)
z <- rnorm(n)
rafalib::mypar()
scatterplot3d(x,y,z, color = group, pch=16, ylab="")
text(8.25, -1.5, label = "y")
abline(v=4, col=3)
```

The line $x=2$ appears to separate the points. But it is actually not the case, which we can see by plotting the data in a couple of two-dimensional points.

```{r pseud-3d-exercise-2, echo=FALSE, fig.height = 3}
rafalib::mypar(1,2)
plot(x,y, col=group, pch =16)
abline(v=2, col=3)
plot(x,z,col=group, pch=16)
abline(v=2, col=3)
```

Why is this happening?

a. Humans are not good at reading pseudo-3D plots.
b. There must be an error in the code.
c. The colors confuse us.
d. Scatterplots should not be used to compare two variables when we have access to 3.

:::
