---
title: "Shrinkage with LASSO and Ridge"
linktitle: "14: Shrinkage with LASSO and Ridge"
read_date: "2022-04-26"
menu:
  content:
    parent: Course content
    weight: 2
type: docs
output:
  blogdown::html_page:
    toc: true
---

## Required Reading

- This page.

### Guiding Questions

- What is *shrinkage*?
- What do we do with too may right-hand-side variables?
- What is LASSO?

# This week is a little different
We will use the lecture slides from my friend and ace economist, Ed Rubin (U of Oregon). They are available [**right here**](https://raw.githack.com/edrubin/EC524W20/master/lecture/005/005-slides.html#1). A couple things first, though:

## The data and packages

Dr. Rubin uses the `ISLR` package's `credit` dataset, which is similar to one we've used before:

```{r credit-data, echo = T}
library(ISLR)
credit = ISLR::Credit
head(credit)

```

We will also need to load the `caret` package (which you're used before), as well as the `glmnet` package, which is new for us.

Now, back to the [**lecture notes for today**](https://raw.githack.com/edrubin/EC524W20/master/lecture/005/005-slides.html#1)



# Try it!

:::fyi
Try it! Use the `wooldridge::wage2` dataset and LASSO to predict `wage`. Since these shrinkage methods (LASSO, ridge) work well with many right-hand-side variables, we can create some additional variables. A good candidate would be squared versions of all of the numeric variables. To do this, we'll use `mutate_if` along with `is.numeric`. The function `mutate_if`, when used in a pipe `%>%`, will check each column to see if the given function is true, then will add a new mutation of that column when true. We need this because we don't want to try to add a squared term for things like factor variables. It will look something like this:

```
data %>%
  mutate_if(is.numeric, list(squared = function(x) x^2))
```


We pass a list with potentially many functions, but only one here in the example. For each passed function, `mutate_if` will create a copy of each existing column, square the data, and call name it by appending the name (here, 'squared') to the original column name.

This will results in some columns we don't want to keep -- `wage_squared` should be dropped since `wage` is the outcome we want to predict. We should also drop `lwage` and `lwage_squared` since those are transformations of our outcome. We want our data to only have the outcome and all the possible predictors so we can simply pass everything but `wage` for our X's and `wage` for our `y`. 

1. Drop all of the copies of the outcome variable and transformations of the outcome variable. Also, drop any of the squared terms of binary variables -- if a variable is $\{0,1\}$, then the squared term is exactly the same. For instance, `married` and `married_squared` are exactly the same.

2. How many predictors (right-side variabes) do we have?

3. Use `glmnet` from the `glmnet` package to run a LASSO (`alpha=1`) using all of the data you've assembled in the `wage2` dataset. Select a wide range of values for `lambda`. You can use `cv.glmnet` to do the cross-validation for you (see the Rubin lecture notes), or you can do it manually as we did with KNN and Regression Trees (test-train split, etc.).

4. Find the `lambda` that minimizes RMSE in the test data. You can extract the optimal lambda from a `cv.glmnet` object by referring to `object$lambda`, and the RMSE using `sqrt(object$cvm)`. These can be used to make the plots in the Rubin lecture notes, and to find the `lambda` that minimizes RMSE. `object$lambda.min` will also tell you the optimal `lambda`.

5. Using the optimal `lambda`, run a final LASSO model.

6. Use `coef` to extract the coefficients from the optimal model. Coefficients that are zeroed out by the LASSO are shown as '.'
- Which variables were "kept" in the model?
- Which variables were eliminated from the model at the optimal `lambda`?
- Is the train RMSE lower than it would be if we just ran OLS with all of the variables?

:::

```{r, eval = F, echo=F}
library(wooldridge)
library(glmnet)

wage2 = wooldridge::wage2 %>%
   mutate_if(is.numeric, list(squared = function(x) x^2)) %>%
  dplyr::select(-wage_squared, -lwage, -lwage_squared, -married_squared, -black_squared, -south_squared, -urban_squared)


lasso_cv = cv.glmnet(
  x = wage2 %>% dplyr::select(-wage) %>% as.matrix(),
  y = wage2$wage,
  standardize = FALSE,
  alpha = 1,
  lambda = 10^seq(from = 1, to = -4, length = 100),
  type.measure = 'mse',
  nfolds = 5
)

## Extract results
lasso_cv_df = data.frame(
  lambda = lasso_cv$lambda,
  rmse = sqrt(lasso_cv$cvm)
)
# Plot
ggplot(
  data = lasso_cv_df,
  aes(x = lambda, y = rmse)
) +
geom_line() +
geom_point(
  data = lasso_cv_df %>% filter(rmse == min(rmse)),
  size = 3.5,
  color = '#18453b'
) +
scale_y_continuous("RMSE") +
scale_x_continuous(
  expression(lambda),
  trans = "log10",
  labels = c("0.1", "10", "1,000", "100,000"),
  breaks = c(0.1, 10, 1000, 100000),
) +
theme_minimal(base_size = 20, base_family = "Fira Sans Book")

lasso_cv_fin = glmnet(
  x = wage2 %>% dplyr::select(-wage) %>% as.matrix(),
  y = wage2$wage,
  standardize = TRUE,
  alpha = 1,
  lambda = lasso_cv$lambda.min
)

```

# Lecture Video
The lecture video for this material is [available here <i class="fas fa-film"></i>](https://mediaspace.msu.edu/media/SSC442+-+Content+14+-+Apr+26th/1_sdivb2nk)