---
title: "Advanced Model Building"
linktitle: "8/9: Linear Regression III"
publishDate: "2021-10-19"
output:
  blogdown::html_page:
    toc: true
menu:
  assignment:
    parent: Labs
    weight: 2
type: docs
weight: 1
editor_options:
  chunk_output_type: console
---

:::fyi

**READ THIS CAREFULLY**

The content below describes both Lab 7 and Lab 8. Lab 7 is Exercise 1; Lab 8 is Exercise 2. Also, you may find some other tasks in the text...

You must turn in a PDF document of your `R Markdown` code. Submit this to D2L by 11:59 PM Eastern Time on Sunday, October 24th for Lab 8; turn in Lab 9 by 11:59 PM Eastern Time on Sunday, October 31st (a very spooky due date!)

:::

## Backstory and Set Up
You still work for Zillow as a junior analyst (sorry). But you're hunting a promotion. Your job is to present some more advanced predictions for housing values in a small geographic area (Ames, IA) using this historical pricing.

As always, let's load the data.

```{r}

Ames <- read.table("https://msudataanalytics.github.io/SSC442/Labs/data/ames.csv",
                 header = TRUE,
                 sep = ",")
```

## Data Cleaning
Oh, the Ames data yet again. It's given us lots of trouble. Many of you have found a few variables (columns) that should be avoided. The main problem is that some columns have only one value in them, or they have only `NA` and one value, so once `lm(...)` drops the `NA` rows, they are left with only one value. Linear regression by OLS does not like variables that don't vary! So, let's be systematic about figuring out which columns in our data are to be avoided.

The `skimr` package is very helpful for seeing what our data contains. Install it, and then use `skim(Ames)` directly in the console (we're just looking at data at the moment). Take a look at the "complete rate" column - this tells us the fraction of observations in that column that are `NA`. If it's very small (see `Alley`), then that variable will be problematic. The "n_unique" column tells us if there are few or many different values - a "1" in "n_unique" is definitely going to be a problem!

You can make a note of those columns that have extremely low "complete rates" and drop them to start off. There are about 5-6 of them that will cause an error if we include them in a regression.

## Linear Models
When exploring linear models in other classes, we often emphasize asymptotic results under distributional assumptions. That is, we make assumptions about the model in order to derive properties of large samples. This general approach is useful for creating and performing hypothesis tests. Frequently, when developing a linear regression model, part of the goal is to **explain** a relationship. However, this isn't always the case. And it's often not a valid approach, as we discussed in this week's content.

So, we will ignore much of what we have learned in other classes (sorry, EC420) and instead simply use regression as a tool to **predict**. Instead of a model which supposedly explains relationships, we seek a model which minimizes **errors**.

To discuss linear models in the context of prediction, we return to the `Ames` data. Accordingly, you should utilize some of the early code from Lab 2 to hasten your progress in this lab.


### Assesing Model Accuracy

There are many metrics to assess the accuracy of a regression model. Most of these measure in some way the average error that the model makes. The metric that we will be most interested in is the root-mean-square error.

$$
\text{RMSE}(\hat{f}, \text{Data}) = \sqrt{\frac{1}{n}\displaystyle\sum_{i = 1}^{n}\left(y_i - \hat{f}(\bf{x}_i)\right)^2}
$$

While for the sake of comparing models, the choice between RMSE and MSE is arbitrary, we have a preference for RMSE, as it has the same units as the response variable. Also, notice that in the prediction context MSE refers to an average, whereas in an ANOVA context, the denominator for MSE may not be $n$.

For a linear model , the estimate of $f$, $\hat{f}$, is given by the fitted regression line.

$$
\hat{y}({\bf{x}_i}) = \hat{f}({\bf{x}_i})
$$

We can write an `R` function that will be useful for performing this calculation.

```{r}
rmse = function(actual, predicted) {
  sqrt(mean((actual - predicted) ^ 2))
}
```

### Model Complexity

Aside from how well a model predicts, we will also be very interested in the complexity (flexibility) of a model. For now, we will only consider nested linear models for simplicity. Then in that case, the more predictors that a model has, the more complex the model. For the sake of assigning a numerical value to the complexity of a linear model, we will use the number of predictors, $p$.

We write a simple `R` function to extract this information from a model.

```{r}
get_complexity = function(model) {
  length(coef(model)) - 1
}
```

When deciding how complex of a model to use, we can utilize two techniques: *forward selection* or *backward selection*. Forward selection means that we start with the simplest model (with a single predictor) and then add one at a time until we decide to stop. Backward selection means that we start with the most complex model (with every available predictor) and then remove one at a time until we decide to stop. There are many criteria for "when to stop". Below, we'll try to give you some intuition on the model-building process.


:::fyi

__EXERCISE 1__

1. Load the `Ames` data. Using `skimr::skim`, find the variables that have a complete rate of below 60\% and drop them. 60\% isn't a magic number by any means, the "right" number is entirely dependent on your data. It is always standard practice to document the fields you have dropped from the data, so make sure you state which variables have been dropped.

2. Take a look at `Utilities`. Use the `table` function to see a tabulation of the values of `Utilities`. Do you see why this field is not likely to be useful to us, or even problematic?

3. Using **forward selection** (that is, select one variable, then select another) create a series of models up to complexity length 15. You may use any variable within the dataset, including categorical variables.

4. Create a chart plotting the model complexity as the $x$-axis variable and RMSE as the $y$-axis variable. Describe any patterns you see. Do you think you should use the full-size model? Why or why not? What criterion are you using to make this statement?

:::


---


### Test-Train Split

There is an issue with fitting a model to all available data then using RMSE to determine how well the model predicts: it is essentially cheating. As a linear model becomes more complex, the RSS, thus RMSE, can never go up. It will only go down---or, in very specific cases where a new predictor is completely uncorrelated with the target, stay the same. This might seem to suggest that in order to predict well, we should use the largest possible model. However, in reality we have fit to a specific dataset, but as soon as we see new data, a large model may (in fact) predict poorly. This is called **overfitting**.

The most common approach to overfitting is to take a dataset of interest and split it in two. One part of the datasets will be used to fit (train) a model, which we will call the **training** data. The remainder of the original data will be used to assess how well the model is predicting, which we will call the **test** data. Test data should *never* be used to train a model---its pupose is to evaluate the fitted model once you've settled on something.[^1]

[^1]: Note that sometimes the terms *evaluation set* and *test set* are used interchangeably. We will give somewhat specific definitions to these later. For now we will simply use a single test set for a training set.

Here we use the `sample()` function to obtain a random sample of the rows of the original data. We then use those row numbers (and remaining row numbers) to split the data accordingly. Notice we used the `set.seed()` function to allow use to reproduce the same random split each time we perform this analysis. Sometimes we don't want to do this; if we want to run lots of independent splits, then we do not need to set the initial seed.

```{r}
set.seed(9)
num_obs = nrow(Ames)

train_index = sample(num_obs, size = trunc(0.50 * num_obs))
train_data = Ames[train_index, ]
test_data = Ames[-train_index, ]
```

We will look at two measures that assess how well a model is predicting: **train RMSE** and **test RMSE**.

$$
\text{RMSE}_\text{Train} = \text{RMSE}(\hat{f}, \text{Train Data}) = \sqrt{\frac{1}{n_{\text{Tr}}}\sum_{i \in \text{Train}}\left(y_i - \hat{f}(\bf{x}_i)\right)^2}
$$

Here $n_{Tr}$ is the number of observations in the train set. Train RMSE will still always go down (or stay the same) as the complexity of a linear model increases. That means train RMSE will not be useful for comparing models, but checking that it decreases is a useful sanity check.

$$
\text{RMSE}_{\text{Test}} = \text{RMSE}(\hat{f}, \text{Test Data}) = \sqrt{\frac{1}{n_{\text{Te}}}\sum_{i \in \text{Test}} \left ( y_i - \hat{f}(\bf{x}_i) \right ) ^2}
$$

Here $n_{Te}$ is the number of observations in the test set. Test RMSE uses the model fit to the training data, but evaluated on the unused test data. This is a measure of how well the fitted model will predict **in general**, not simply how well it fits data used to train the model, as is the case with train RMSE. What happens to test RMSE as the size of the model increases? That is what we will investigate.

We will start with the simplest possible linear model, that is, a model with no predictors.

```{r}
fit_0 = lm(SalePrice ~ 1, data = train_data)
get_complexity(fit_0)

# train RMSE
sqrt(mean((train_data$SalePrice - predict(fit_0, train_data)) ^ 2))
# test RMSE
sqrt(mean((test_data$SalePrice - predict(fit_0, test_data)) ^ 2))
```

The previous two operations obtain the train and test RMSE. Since these are operations we are about to use repeatedly, we should use the function that we happen to have already written.

```{r}
# train RMSE
rmse(actual = train_data$SalePrice, predicted = predict(fit_0, train_data))
# test RMSE
rmse(actual = test_data$SalePrice, predicted = predict(fit_0, test_data))
```

This function can actually be improved for the inputs that we are using. We would like to obtain train and test RMSE for a fitted model, given a train or test dataset, and the appropriate response variable.

```{r}
get_rmse = function(model, data, response) {
  rmse(actual = subset(data, select = response, drop = TRUE),
       predicted = predict(model, data))
}
```

By using this function, our code becomes easier to read, and it is more obvious what task we are accomplishing.

```{r}
get_rmse(model = fit_0, data = train_data, response = "SalePrice") # train RMSE
get_rmse(model = fit_0, data = test_data, response = "SalePrice") # test RMSE
```

**Try it:** Apply this basic function with different arguments. Do you understand how we've nested functions within functions?

**Try it:** Define a total of five models using the first five models you fit in Exercise 1. Define these as `fit_1` through `fit_5`


### Adding Flexibility to Linear Models

Each successive model we fit will be more and more flexible using both interactions and polynomial terms. We will see the training error decrease each time the model is made more flexible. We expect the test error to decrease a number of times, then eventually start going up, as a result of overfitting. To better understand the relationship between train RMSE, test RMSE, and model complexity, we'll explore the results from Exercise 1.

Hopefully, you tried the in-line excercise above. If so, we can create a list of the models fit.

```{r, eval=FALSE}
model_list = list(fit_1, fit_2, fit_3, fit_4, fit_5)
```

We then obtain train RMSE, test RMSE, and model complexity for each. In doing so, we'll introduce a handy function from `R` called `sapply()`. You can likely intuit what it does by looking at the code below.

```{r, eval=FALSE}
train_rmse = sapply(model_list, get_rmse, data = train_data, response = "SalePrice")
test_rmse = sapply(model_list, get_rmse, data = test_data, response = "SalePrice")
model_complexity = sapply(model_list, get_complexity)
```

**Try it:** Run `?sapply()` to understand what are valid arguments to the function.

Once you've done this, you'll notice the following:

```{r, eval=FALSE}
# This is the same as the apply command above

test_rmse = c(get_rmse(fit_1, test_data, "SalePrice"),
              get_rmse(fit_2, test_data, "SalePrice"),
              get_rmse(fit_3, test_data, "SalePrice"),
              get_rmse(fit_4, test_data, "SalePrice"),
              get_rmse(fit_5, test_data, "SalePrice"))
```

We can plot the results. If you execute the code below, you'll see the train RMSE in blue, while the test RMSE is given in orange.[^3]

[^3]: The train RMSE is guaranteed to follow this non-increasing pattern. The same is not true of test RMSE. We often see a nice U-shaped curve. There are theoretical reasons why we should expect this, but that is on average. Because of the randomness of one test-train split, we may not always see this result. Re-perform this analysis with a different seed value and the pattern may not hold. We will discuss why we expect this next chapter. We will discuss how we can help create this U-shape much later. Also, we might intuitively expect train RMSE to be lower than test RMSE. Again, due to the randomness of the split, you may get (un)lucky and this will not be true.

```{r, eval=FALSE}
plot(model_complexity, train_rmse, type = "b",
     ylim = c(min(c(train_rmse, test_rmse)) - 0.02,
              max(c(train_rmse, test_rmse)) + 0.02),
     col = "dodgerblue",
     xlab = "Model Size",
     ylab = "RMSE")
lines(model_complexity, test_rmse, type = "b", col = "darkorange")
```

We could also summarize the results as a table. `fit_1` is the least flexible, and `fit_5` is the most flexible. If we were to do this (see the exercise below) we would see that Train RMSE decreases as flexibility increases forever. However, this may not be the case for the Test RMSE.

| Model   | Train RMSE        | Test RMSE        | Predictors              |
|---------|-------------------|------------------|-------------------------|
| `fit_1` | RMSE$_{\text{train}}$ for model 1 | RMSE$_{\text{test}}$ for model 1 | put predictors here|
| ...| ... | .... | ... |
| `fit_5` | RMSE$_{\text{train}}$ for model 5  | RMSE$_{\text{train}}$ for model 5  | $p$ predictors |


To summarize:

- **Underfitting models:** In general *High* Train RMSE, *High* Test RMSE.
- **Overfitting models:** In general *Low* Train RMSE, *High* Test RMSE.

Specifically, we say that a model is overfitting if there exists a less complex model with lower Test RMSE.[^2] Then a model is underfitting if there exists a more complex model with lower Test RMSE.

[^2]: The labels of under and overfitting are *relative* to the best model we see. Any model more complex with higher Test RMSE is overfitting. Any model less complex with higher Test RMSE is underfitting.

:::fyi

__EXERCISE 2__

(AKA Lab 8)

1. Make a table exactly like the table above for the 15 models you fit in Exercise 1.

2. **This question should be the most time-consuming question.** Using any method you choose and any number of regressors, predict `SalePrice`. Calculate the Train and Test RMSE. Your goal is to have a lower Test RMSE than others in the class.

3. In a short paragraph, describe the resulting model. Discuss how you arrived at this model, what interactions you're using (if any) and how confident you are that your prediction will perform well, relative to other people in the class.

4. **Difficult; extra credit:** Visualize your final model in a sensible way and provide a two-paragraph interpretation.

:::

A final note on the analysis performed here; we paid no attention whatsoever to the "assumptions" of a linear model. We only sought a model that **predicted** well, and paid no attention to a model for **explaination**. Hypothesis testing did not play a role in deciding the model, only prediction accuracy. Collinearity? We don't care. Assumptions? Still don't care. Diagnostics? Never heard of them. (These statements are a little over the top, and not completely true, but just to drive home the point that we only care about prediction. Often we latch onto methods that we have seen before, even when they are not needed.)
