---
title: "ggplot2: Everything you ever wanted to know"
linktitle: "2: Visualization II"
read_date: "2021-02-04"
draft: false
publishDate: "2021-01-25"
lastmod: "`r Sys.Date()`"
output:
  blogdown::html_page:
    toc: true
menu:
  example:
    parent: Examples
    weight: 1
type: docs
weight: 1
editor_options:
  chunk_output_type: console
---

## Add-on packages {#add-on-packages}

The power of __ggplot2__ is augmented further due to the availability of add-on packages.
The remaining changes needed to put the finishing touches on our plot require the __ggthemes__ and __ggrepel__ packages.

The style of a __ggplot2__ graph can be changed using the `theme` functions. Several themes are included as part of the __ggplot2__ package. In fact, for most of the plots in this book, we use a function in the __dslabs__ package that automatically sets a default theme:

```{r, eval}
ds_theme_set()
```

Many other themes are added by the package __ggthemes__. Among those are the `theme_economist` theme that we used. After installing the package, you can change the style by adding a layer like this:

```{r, eval = FALSE}
library(ggthemes)
p + theme_economist()
```

You can see how some of the other themes look by simply changing the function. For instance, you might try the `theme_fivethirtyeight()` theme instead.

The final difference has to do with the position of the labels. In our plot, some of the labels fall on top of each other. The add-on package __ggrepel__ includes a geometry that adds labels while ensuring that they don't fall on top of each other. We simply change `geom_text` with `geom_text_repel`.

## Putting it all together

Now that we are done testing, we can write one piece of code that produces our desired plot from scratch.


```{r final-ggplot-example}
library(ggthemes)
library(ggrepel)

r <- murders %>%
  summarize(rate = sum(total) /  sum(population) * 10^6) %>%
  pull(rate)

murders %>% ggplot(aes(population/10^6, total, label = abb)) +
  geom_abline(intercept = log10(r), lty = 2, color = "darkgrey") +
  geom_point(aes(col=region), size = 3) +
  geom_text_repel() +
  scale_x_log10() +
  scale_y_log10() +
  xlab("Populations in millions (log scale)") +
  ylab("Total number of murders (log scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region") +
  theme_economist_white()
```

```{r, echo=FALSE}
ds_theme_set()
```


## Quick plots with `qplot` {#qplot}

We have learned the powerful approach to generating visualization with ggplot. However, there are instances in which all we want is to make a quick plot of, for example, a histogram of the values in a vector, a scatterplot of the values in two vectors, or a boxplot using categorical and numeric vectors. We demonstrated how to generate these plots with `hist`, `plot`, and `boxplot`. However, if we want to keep consistent with the ggplot style, we can use the function `qplot`.

If we have values in two vectors, say:

```{r}
data(murders)
x <- log10(murders$population)
y <- murders$total
```

and we want to make a scatterplot with ggplot, we would have to type something like:

```{r, eval=FALSE}
data.frame(x = x, y = y) %>%
  ggplot(aes(x, y)) +
  geom_point()
```

This seems like too much code for such a simple plot.
The `qplot` function sacrifices the flexibility provided by the `ggplot` approach, but allows us to generate a plot quickly.

```{r qplot-example-0, eval = FALSE}
qplot(x, y)
```

Although we won't discuss `qtplot` in much detail, you should feel free to use it in the early stages of your data exploration. Once you're settled on a final design, then move to `ggplot`.

## Grids of plots

There are often reasons to graph plots next to each other. The __gridExtra__ package permits us to do that:

```{r gridExtra-example, warning=FALSE, message=FALSE, fig.height=2.5, fig.width=5}
library(gridExtra)
p1 <- qplot(x)
p2 <- qplot(x,y)
grid.arrange(p1, p2, ncol = 2)
```


## In-class exercises
If we have time, let's explore some new geom's.

I'll break you out into Breakout Rooms of 4-5 people each. In your room, I want you to explore one of the following 5 `ggplot` geoms. Use ?geom_name to learn about the *aes* arguments each takes / requires.

For whichever geom your group chooses, use the `dslabs::gapminder` data 

```
library(dslabs)
gapminder = dslabs::gapminder
```



__geom_density_2d_filled()__

- This is a 2d density plot which shows correlation between two variables. Try `infant_mortality` and GDP per capita

- __Challenge__: Show infant mortality's relationship to gdp per capita in the year 2007

- x = infant_mortality, y = calculate gdp per capita on the fly (in the ggplot call)

__geom_tile()__

- This is a heatmap type plot

- __Challenge__: Show infant_mortality by year for each country in continent=='Americas'

- Hint: x = year, y = country

__geom_smooth()__

- This is plot that shows the mean and variance (standard deviation) over time

- __Challenge__: Show the distribution of infant mortality across the world for 1960 and 2007, with a separate density for each of those two years

- Hint: x = infant_mortality, "+ facet_wrap"


__geom_qq()__

- This plot helps us tell if a variable's distribution is close to normal

- __Challenge__: Show a qq plot of fertility in 2007 across the world. Is it normally distributed? 

- Hint: This one takes a new `aes(...)` argument: sample = fertility

- Hint: geom_qq and geom_qq_line do two different things, but can be combined. geom_qq_line is a lot like geom_abline()



__geom_rug()__

- This plot shows both the point scatter and the density of each variable. `geom_rug` adds density in the margins, use with geom_point

- __Challenge__:Visualize the relationship between gdp per capita and infant mortality that also communicates the density of each variable

- Hint: x = gdp per capita (calculate in ggplot call), y = infant_mortality

- Hint: use alpha = .1 to make rug lines show overlap


 
:::fyi

__TRY IT__

Start by loading the __dplyr__ and __ggplot2__ library as well as the `murders` and `heights` data.

```{r}
library(dplyr)
library(ggplot2)
library(dslabs)
data(heights)
data(murders)
```

1. With __ggplot2__ plots can be saved as objects. For example we can associate a dataset with a plot object like this

```{r, eval = FALSE}
p <- ggplot(data = murders)
```

Because `data` is the first argument we don't need to spell it out

```{r, eval = FALSE}
p <- ggplot(murders)
```

and we can also use the pipe:

```{r, eval = FALSE}
p <- murders %>% ggplot()
```

What is class of the object `p`?


2. Remember that to print an object you can use the command `print` or simply type the object.
Print the object `p` defined in exercise one and describe what you see.

a. Nothing happens.
b. A blank slate plot.
c. A scatterplot.
d. A histogram.


3. Using the pipe `%>%`, create an object `p` but this time associated with the `heights` dataset instead of the `murders` dataset.


4. What is the class of the object `p` you have just created?


5. Now we are going to add a layer and the corresponding aesthetic mappings. For the murders data we plotted total murders versus population sizes. Explore the `murders` data frame to remind yourself what are the names for these two variables and select the correct answer. __Hint__: Look at `?murders`.

a. `state` and `abb`.
b. `total_murders` and `population_size`.
c. `total` and `population`.
d. `murders` and `size`.

6. To create the scatterplot we add a layer with `geom_point`. The aesthetic mappings require us to define the x-axis and y-axis variables, respectively. So the code looks like this:

```{r, eval=FALSE}
murders %>% ggplot(aes(x = , y = )) +
  geom_point()
```

except we have to define the two variables `x` and `y`. Fill this out with the correct variable names.


7. Note that if we don't use argument names, we can obtain the same plot by making sure we enter the variable names in the right order like this:

```{r, eval=FALSE}
murders %>% ggplot(aes(population, total)) +
  geom_point()
```

Remake the plot but now with total in the x-axis and population in the y-axis.


8. If instead of points we want to add text, we can use the `geom_text()` or `geom_label()` geometries. The following code

```{r, eval=FALSE}
murders %>% ggplot(aes(population, total)) + geom_label()
```

will give us the error message: `Error: geom_label requires the following missing aesthetics: label`

Why is this?

a. We need to map a character to each point through the label argument in aes.
b. We need to let `geom_label` know what character to use in the plot.
c. The `geom_label` geometry does not require  x-axis and y-axis values.
d. `geom_label` is not a ggplot2 command.


9. Rewrite the code above to use abbreviation as the label through `aes`


10. Change the color of the labels to blue. How will we do this?

a. Adding a column called `blue` to `murders`.
b. Because each label needs a different color we map the colors through `aes`.
c. Use the `color` argument in `ggplot`.
d. Because we want all colors to be blue, we do not need to map colors, just use the color argument in `geom_label`.


11. Rewrite the code above to make the labels blue.


12. Now suppose we want to use color to represent the different regions. In this case which of the following is most appropriate:

a. Adding a column called `color` to `murders` with the color we want to use.
b. Because each label needs a different color we map the colors through the color argument of `aes` .
c. Use the `color` argument in `ggplot`.
d. Because we want all colors to be blue, we do not need to map colors, just use the color argument in `geom_label`.


13. Rewrite the code above to make the labels' color be determined by the state's region.

14. Now we are going to change the x-axis to a log scale to account for the fact the distribution of population is skewed. Let's start by defining an object `p` holding the plot we have made up to now

```{r, eval=FALSE}
p <- murders %>%
  ggplot(aes(population, total, label = abb, color = region)) +
  geom_label()
```

To change the y-axis to a log scale we learned about the `scale_x_log10()` function. Add this layer to the object `p` to change the scale and render the plot.

15. Repeat the previous exercise but now change both axes to be in the log scale.

16. Now edit the code above to add the title "Gun murder data" to the plot. Hint: use the `ggtitle` function.
:::
