---
title: "Advanced Model Building"
linktitle: "7/8: Linear Regression III"
output:
  blogdown::html_page:
    toc: true
menu:
  assignment:
    parent: Labs
    weight: 2
type: docs
weight: 1
editor_options:
  chunk_output_type: console
---


:::fyi

**READ THIS CAREFULLY**

The content below describes both Lab 7 and Lab 8. Lab 7 is Exercise 1; Lab 8 is Exercise 2.

You must turn in a PDF document of your `R Markdown` code. Submit this to D2L by 11:59 PM Eastern Time on Monday, March 27th for Lab 7. 

Lab 8 is due Monday, April 3rd. 

Submit labs separately. Uploading both labs to Lab 7 may result in a 0 on Lab 8.

:::

## Backstory and Set Up
You still work for Zillow as a junior analyst (sorry). But you're hunting a promotion. Your job is to present some more advanced predictions for housing values in a small geographic area (Ames, IA) using this historical pricing.

As always, let's load the data.

```{r}

Ames <- read.table('https://www.openintro.org/data/csv/ames.csv', 
                   header = TRUE,
                   sep = ',') 
```

## Data Cleaning
**Do not skip this section. This isn't your kitchen junk drawer -- you can't get away with not cleaning your data.**

Oh, the Ames data yet again. 

We're going to use the *unpolished* version of this data that was the original dataset. This is good (a little more data) and bad (slight changes in names of data fields, lots of missing data).

The `skimr` package is very helpful for seeing what our data contains. Install it, and then use `skim(Ames)` directly in the console (we're just looking at data at the moment -- do **not** put `skim` output into your RMarkdwon output - it will give you an error). Take a look at the "complete rate" column - this tells us the fraction of observations in that column that are `NA`. If it's very small (see `Alley`), then that variable will be problematic. The "n_unique" column tells us if there are few or many different values - a "1" in "n_unique" is definitely going to be a problem and you must drop that variable.

You can make a note of those columns that have extremely low "complete rates" and drop them to start off. There are about 5-6 of them that will cause an error if we include them in a regression.

### Many models do not like `NA` values
`predict` has some unusual behavior that can give unexpected results. Thus far, we have mostly used `predict(myOLS)`, which gives the predicted values from a model *using the same data that estimated the model*. When we ask `lm` (or, later, other machine learning models) to estimate a model, it will drop any rows of our data that contain a `NA` value for any of the variables used in the estimation. If your regression is `price ~ GrLivArea`, then it will check `price` and `GrLivArea` for NA's. If you add another variable, then you add another possible set of `NA` values that can be dropped, and R will estimate the model on a subset of the data.

This will mess with your measure of $RMSE$ - if you compare two models that use different sets of the data, then you aren't really comparing the fit very well. Because of that, we need to take a moment to check our data.

Ames has a lot of variables, and in this assignment, you're going to be asked to construct 15 regressions of increasing complexity. So we're going to choose 15 variables to be explanatory, one variable we want to be the target variable we predict (your choice for Lab 7), plus `price`, which will be our target variable for Lab 8. Make sure we have dropped any observation (row) that contains one or more `NA` value across the 17 columns. You should have already dropped any variable with a very low complete rate or small "n_unique" count. 

Make a character vector of the variables you think will best predict your chosen variable. Then, make a new version of `Ames` that contains only those 17 variables (you can use `dplyr::select` or any other method). Once you've done that, use `na.omit` to make a new, clean version of Ames that has (1) no NA's in it, and (2) 16 variables of your choice (as long as one is `price`). This way, we know every model we make will have the same number of rows in it as none will be dropped due to NA values.



## Linear Models
When exploring linear models in other classes, we often emphasize asymptotic results under distributional assumptions. That is, we make assumptions about the model in order to derive properties of large samples. This general approach is useful for creating and performing hypothesis tests. Frequently, when developing a linear regression model, part of the goal is to **explain** a relationship. However, this isn't always the case. And it's often not a valid approach, as we discussed in this week's content.

So, we will ignore much of what we have learned in other classes (sorry, EC420) and instead simply use regression as a tool to **predict**. Instead of a model which supposedly explains relationships, we seek a model which minimizes **errors**.

To discuss linear models in the context of prediction, we return to the `Ames` data. Accordingly, you should utilize some of the early code from Lab 2 to hasten your progress in this lab.


### Assesing Model Accuracy

There are many metrics to assess the accuracy of a regression model. Most of these measure in some way the average error that the model makes. The metric that we will be most interested in is the root-mean-square error.

$$
\text{RMSE}(\hat{f}, \text{Data}) = \sqrt{\frac{1}{n}\displaystyle\sum_{i = 1}^{n}\left(y_i - \hat{f}(\bf{x}_i)\right)^2}
$$

While for the sake of comparing models, the choice between RMSE and MSE is arbitrary, we have a preference for RMSE, as it has the same units as the response variable. Also, notice that in the prediction context MSE refers to an average, whereas in an ANOVA context, the denominator for MSE may not be $n$.

For a linear model , the estimate of $f$, $\hat{f}$, is given by the fitted regression line.

$$
\hat{y}({\bf{x}_i}) = \hat{f}({\bf{x}_i})
$$

We can write an `R` function that will be useful for performing this calculation.

```{r}
rmse = function(actual, predicted) {
  sqrt(mean((actual - predicted) ^ 2))
}
```

### Model Complexity

Aside from how well a model predicts, we will also be very interested in the complexity (flexibility) of a model. For now, we will only consider nested linear models for simplicity. Then in that case, the more predictors that a model has, the more complex the model. For the sake of assigning a numerical value to the complexity of a linear model, we will use the number of predictors, $p$.

We write a simple `R` function to extract this information from a model.

```{r}
get_complexity = function(model) {
  length(coef(model)) - 1
}
```

When deciding how complex of a model to use, we can utilize two techniques: *forward selection* or *backward selection*. Forward selection means that we start with the simplest model (with a single predictor) and then add one at a time until we decide to stop. Backward selection means that we start with the most complex model (with every available predictor) and then remove one at a time until we decide to stop. There are many criteria for "when to stop". Below, we'll try to give you some intuition on the model-building process.


:::fyi

__EXERCISE 1__

1. Load the `Ames` data. Take a look at `Utilities`. Use the `table` function to see a tabulation of the values of `Utilities`. Do you see why this field is not likely to be useful to us, or even problematic?

2. Using `skimr::skim`, find the variables that have a complete rate of below 60\% and drop them. 60\% isn't a magic number by any means, the "right" number is entirely dependent on your data. It is always standard practice to document the fields you have dropped from the data, so make sure you state which variables have been dropped. Also using `skim`, note the variables with values for "n_unique" equal to 1 and drop them. 

3. Then, choose one numeric variable to be your target variable (the variable you want to explain or predict), and the 15 variables you want to use to predict it - they can be numeric or categorical, both will work. Then, use `na.omit` to make a clean version of `Ames` where every variable is 100% complete. See the instructions above under "Data Cleaning". Include `price` in your cleaning, but don't use it yet -- it will be the target variable for your Lab 8.


4. Using **forward selection** (that is, select one variable, then select another) create a series of models up to complexity length 15. While you can code each of the 15 regressions separately, if you really want to be efficient, try to use a loop along with a character vector of your 15 variables to "step" through the 15 regressions. Use a list object to hold your results, and use `lapply` along with your RMSE function(s) to get a list of the RMSE's, one for each model.

5. Make a `data.frame` or `tibble` of the RMSE results and the model complexity (the function `unlist` is helpful when you have a list of identical types, as should be the case with your measures of model complexity and RMSE). Create a chart plotting the model complexity as the $x$-axis variable and RMSE as the $y$-axis variable. Describe any patterns you see. Do you think you should use the full-size model? Why or why not? What criterion are you using to make this statement?

:::


---


### Test-Train Split (Lab 8)

Here, we will use `price` as our target variable. You can re-use your code for cleaning from above (you included `price` there, right?), and if you want to choose different predictors, feel free to do so.


There is an issue with fitting a model to all available data then using RMSE to determine how well the model predicts: it is essentially cheating. As a linear model becomes more complex, the RSS, thus RMSE, can never go up. It will only go down---or, in very specific cases where a new predictor is completely uncorrelated with the target, stay the same. This might seem to suggest that in order to predict well, we should use the largest possible model. However, in reality we have fit to a specific dataset, but as soon as we see new data, a large model may (in fact) predict poorly. This is called **overfitting**.

The most common approach to overfitting is to take a dataset of interest and split it in two. One part of the datasets will be used to fit (train) a model, which we will call the **training** data. The remainder of the original data will be used to assess how well the model is predicting, which we will call the **test** data. Test data should *never* be used to train a model---its pupose is to evaluate the fitted model once you've settled on something.[^1]

[^1]: Note that sometimes the terms *evaluation set* and *test set* are used interchangeably. We will give somewhat specific definitions to these later. For now we will simply use a single test set for a training set.

Here we use the `sample()` function to obtain a random sample of the rows of the original data. We then use those row numbers (and remaining row numbers) to split the data accordingly. Notice we used the `set.seed()` function to allow use to reproduce the same random split each time we perform this analysis. Sometimes we don't want to do this; if we want to run lots of independent splits, then we do not need to set the initial seed.

```{r}
set.seed(9)
num_obs = nrow(Ames)

train_index = sample(num_obs, size = trunc(0.50 * num_obs))
train_data = Ames[train_index, ]
test_data = Ames[-train_index, ]
```

We will look at two measures that assess how well a model is predicting: **train RMSE** and **test RMSE**.

$$
\text{RMSE}_\text{Train} = \text{RMSE}(\hat{f}, \text{Train Data}) = \sqrt{\frac{1}{n_{\text{Tr}}}\sum_{i \in \text{Train}}\left(y_i - \hat{f}(\bf{x}_i)\right)^2}
$$

Here $n_{Tr}$ is the number of observations in the train set. Train RMSE will still always go down (or stay the same) as the complexity of a linear model increases. That means train RMSE will not be useful for comparing models, but checking that it decreases is a useful sanity check.

$$
\text{RMSE}_{\text{Test}} = \text{RMSE}(\hat{f}, \text{Test Data}) = \sqrt{\frac{1}{n_{\text{Te}}}\sum_{i \in \text{Test}} \left ( y_i - \hat{f}(\bf{x}_i) \right ) ^2}
$$

Here $n_{Te}$ is the number of observations in the test set. Test RMSE uses the model fit to the training data, but evaluated on the unused test data. This is a measure of how well the fitted model will predict **in general**, not simply how well it fits data used to train the model, as is the case with train RMSE. What happens to test RMSE as the size of the model increases? That is what we will investigate.

We will start with the simplest possible linear model, that is, a model with no predictors.

```{r}
fit_0 = lm(price ~ 1, data = train_data)
get_complexity(fit_0)

# train RMSE
sqrt(mean((train_data$price - predict(fit_0, train_data)) ^ 2))
# test RMSE
sqrt(mean((test_data$price - predict(fit_0, test_data)) ^ 2))
```

The previous two operations obtain the train and test RMSE. Since these are operations we are about to use repeatedly, we should use the function that we happen to have already written.

```{r}
# train RMSE
rmse(actual = train_data$price, predicted = predict(fit_0, train_data))
# test RMSE
rmse(actual = test_data$price, predicted = predict(fit_0, test_data))
```

This function can actually be improved for the inputs that we are using. We would like to obtain train and test RMSE for a fitted model, given a train or test dataset, and the appropriate response variable.

```{r}
get_rmse = function(model, data, response) {
  rmse(actual = subset(data, select = response, drop = TRUE),
       predicted = predict(model, data))
}
```

By using this function, our code becomes easier to read, and it is more obvious what task we are accomplishing.

```{r}
get_rmse(model = fit_0, data = train_data, response = "price") # train RMSE
get_rmse(model = fit_0, data = test_data, response = "price") # test RMSE
```

**Try it:** Apply this basic function with different arguments. Do you understand how we've nested functions within functions?

**Try it:** Define a total of **five** models with increasing complexity (see Exercises 2, Question 1)


### Adding Flexibility to Linear Models

Each successive model we fit will be more and more flexible using both interactions and polynomial terms. We will see the training error decrease each time the model is made more flexible. We expect the test error to decrease a number of times, then eventually start going up, as a result of overfitting. To better understand the relationship between train RMSE, test RMSE, and model complexity, we'll explore the results from Exercise 1.

Hopefully, you tried the in-line excercise above. If so, we can create a list of the models fit.

```{r, eval=FALSE}
model_list = list(fit_1, fit_2, fit_3, fit_4, fit_5)
```

We then obtain train RMSE, test RMSE, and model complexity for each. In doing so, we'll introduce a handy function from `R` called `sapply()`. You can likely intuit what it does by looking at the code below.

```{r, eval=FALSE}
train_rmse = sapply(model_list, get_rmse, data = train_data, response = "price")
test_rmse = sapply(model_list, get_rmse, data = test_data, response = "price")
model_complexity = sapply(model_list, get_complexity)
```

**Try it:** Run `?sapply()` to understand what are valid arguments to the function. It's just like `lapply`, but instead of outputting a list, it outputs a vector of results.

Once you've done this, you'll notice the following:

```{r, eval=FALSE}
# This is the same as the apply command above

test_rmse = c(get_rmse(fit_1, test_data, "price"),
              get_rmse(fit_2, test_data, "price"),
              get_rmse(fit_3, test_data, "price"),
              get_rmse(fit_4, test_data, "price"),
              get_rmse(fit_5, test_data, "price"))
```

We can plot the results. If you execute the code below, you'll see the train RMSE in blue, while the test RMSE is given in orange.[^3]

[^3]: The train RMSE is guaranteed to follow this non-increasing pattern as long as no data is being dropped when new variables are added (see [Data Cleaning](#Data Cleaning) above). The same is not true of test RMSE. We often see a nice U-shaped curve. There are theoretical reasons why we should expect this, but that is on average. Because of the randomness of one test-train split, we may not always see this result. Re-perform this analysis with a different seed value and the pattern may not hold. We will discuss why we expect this next chapter. We will discuss how we can help create this U-shape much later. Also, we might intuitively expect train RMSE to be lower than test RMSE. Again, due to the randomness of the split, you may get (un)lucky and this will not be true.

```{r, eval=FALSE}
plot(model_complexity, train_rmse, type = "b",
     ylim = c(min(c(train_rmse, test_rmse)) - 0.02,
              max(c(train_rmse, test_rmse)) + 0.02),
     col = "dodgerblue",
     xlab = "Model Size",
     ylab = "RMSE")
lines(model_complexity, test_rmse, type = "b", col = "darkorange")
```

We could also summarize the results as a table. `fit_1` is the least flexible, and `fit_5` is the most flexible. If we were to do this (see the exercise below) we would see that Train RMSE decreases as flexibility increases forever. However, this may not be the case for the Test RMSE.

| Model   | Train RMSE        | Test RMSE        | Predictors              |
|---------|-------------------|------------------|-------------------------|
| `fit_1` | RMSE$_{\text{train}}$ for model 1 | RMSE$_{\text{test}}$ for model 1 | put predictors here|
| ...| ... | .... | ... |
| `fit_5` | RMSE$_{\text{train}}$ for model 5  | RMSE$_{\text{train}}$ for model 5  | $p$ predictors |


To summarize:

- **Underfitting models:** In general *High* Train RMSE, *High* Test RMSE.
- **Overfitting models:** In general *Low* Train RMSE, *High* Test RMSE.

Specifically, we say that a model is overfitting if there exists a less complex model with lower Test RMSE.[^2] Then a model is underfitting if there exists a more complex model with lower Test RMSE.

[^2]: The labels of under and overfitting are *relative* to the best model we see. Any model more complex with higher Test RMSE is overfitting. Any model less complex with higher Test RMSE is underfitting.

:::fyi

__EXERCISE 2__

(AKA Lab 8)

1. Using any method you choose and any number of regressors, predict `price` with no fewer than five models of increasing complexity. Complexity can be increased by adding variables *or* by adding interactions of existing variables. Put the five models into a list.

2. Calculate the Train and Test RMSE. Your goal is to have a lower *Test* RMSE than others in the class. 

2. Make a table exactly like the table above for the 5 models you just fit. The first column should have the name of the model (e.g. `fit_1`). Hint: you can get the names of the entries in a list using `names(model_list)` provided you named the list items when you added them.

3. In a short paragraph, describe the resulting model. Discuss how you arrived at this model, what interactions you're using (if any) and how confident you are that your prediction will perform well, relative to other people in the class.

4. **Difficult; extra credit:** Visualize your final model in a sensible way and provide a two-paragraph interpretation.

:::

A final note on the analysis performed here; we paid no attention whatsoever to the "assumptions" of a linear model. We only sought a model that **predicted** well, and paid no attention to a model for **explaination**. Hypothesis testing did not play a role in deciding the model, only prediction accuracy. Collinearity? We don't care. Assumptions? Still don't care. Diagnostics? Never heard of them. (These statements are a little over the top, and not completely true, but just to drive home the point that we only care about prediction. Often we latch onto methods that we have seen before, even when they are not needed.)
